---
title: "Lab10"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy_file
```

```{r}
candy <- read.csv(candy_file, row.names=1)
head(candy)
```
##Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 different candy types in this dataset. 

##Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```

There are 38 fruity candy types in the dataset. 

```{r}
candy["Twix", ]$winpercent
```

##Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Snickers", ]$winpercent
```

My favorite candy in the dataset is Snickers. Snickers has a winpercent value of 76.67378. 

##Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

The winpercent value for Kit Kat is 76.7686. 

##Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

The winpercent value for Tootsie Roll Snack Bars is 49.6535. 

```{r}
library("skimr")
skim(candy)
```

##Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset? 
Yes, variable sd looks to be on a different scale to the majority of the other columns in the dataset.

##Q7. What do you think a zero and one represent for the candy$chocolate column?

The zero for the chocolate column represents false (that it is NOT chocolate). The one for the chocolate column represe
```{r}
candy
```

##Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent) 
```
##Q9. Is the distribution of winpercent values symmetrical?
Based off the histogram, the distribution of winpercent values are not symmetrical.

##Q10. Is the center of the distribution above or below 50%?
The center of the distribution is below 50%. 

##Q11. On average is chocolate candy higher or lower ranked than fruit candy? ----
On average chocolate candy (60.92153) is higher ranked than fruit candy(44.11974). 

```{r}
chocolate <- candy$winpercent[as.logical(candy$chocolate)]
fruity <- candy$winpercent[as.logical(candy$fruity)]
```

```{r}
mean(chocolate)
mean(fruity)
```

##Q12. Is this difference statistically significant?
The difference is statistically significant because the p-value (2.871e-08) is less than 0.05 meaning the null hypothesis is rejected.
```{r}
t.test(chocolate, fruity)
```

##Q13. What are the five least liked candy types in this set?
The five least liked candies in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters. 

```{r}
head(candy[order(candy$winpercent),], n=5)

```


##Q14. What are the top 5 all time favorite candy types out of this set?
The five top candies in this set are snickers, Kit Kat, Twix, ReeseÕs Miniatures and ReeseÕs Peanut Butter cup.
```{r}
tail(candy[order(candy$winpercent),], n=5)
```
 

##Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library("tidyverse")
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```

##Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

##Q17. What is the worst ranked chocolate candy?
The worst ranked chocolate candy is sixlets. 

##Q18. What is the best ranked fruity candy?
The best ranked fruity candy is Starburst. 

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

```{r}
library(ggrepel)
```


```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
##Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck? 

Reeses Miniatures is the candy with the highest winpercent and the lowest pricepercent. 

##Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
The top 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring pop, Hersheys Krackel and Hersheys Milk Chocolate. Out of these 5 expensive candies, the least popular candy is Nik L Nip. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```
##Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
library(corrplot)
```
##Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
The two variables that are anti-correlated are fruity and chocolate. 

##Q23. Similarly, what two variables are most positively correlated?
The two variables that are most positively correlated are chocolate and bar 

```{r}
cij <- cor(candy)
corrplot(cij)
```
###CHECK my PCA***********************
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
ggplotly(p)
```

##Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus are the variables that are picked up strongly by PC1 in the positive direction. Yes, this makes sense to me because the scatterplot created from PC1 vs PC2 results correlate with the bar graph generated. The fruity candy was found to be positive in the scatterplot PC1 vs PC2. 

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```





