---
title: "Lab 19"
author: "Karen Guerrero"
format: pdf
---

```{r}
##install.packages("datapasta")
```


## Is Pertussis on the rise

Scrape the CDC data on Pertussis cases per year in the US from their website here: 

https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html
                                                                                                             
```{r}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L),
         No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617)
       )
```

```{r}
View(cdc)
```

```{r}
library(ggplot2)
```

##Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
ggplot(cdc) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x="year", y="No.Reported.Pertussis.Cases", title="Pertussis Cases by Year(1922-2019)")
```


```{r}
baseplot <- ggplot(cdc) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x="year", y="No.Reported.Pertussis.Cases", title="Pertussis Cases by Year(1922-2019)")
```

##Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?
```{r}
ggplot(cdc) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() + geom_vline(xintercept=1996) + geom_vline(xintercept=1946)
  labs(x="year", y="No.Reported.Pertussis.Cases", title="Pertussis Cases by Year(1922-2019)")
```


There are two main Pertussis vaccine types, so-called whole-cell Pertusis (wP) and acellular Pertussis vaccines (aP). The older wP was introduced in 1946 and was highly effective. The switch to aP was in 1996. 

```{r}
baseplot + geom_vline(xintercept = 1946, color="red", linetype=2) + geom_vline(xintercept = 1996, color="blue", linetype=2) + geom_text(x=1948, y=250000, label="wP", col="red") + geom_text(x=1998, y=250000, label="aP", col="blue")
```
##Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of aP vaccine the number of reported Pertussis cases began to slightly increase. I think introducing a new type of Pertussis vaccine created a fear towards the new type of vaccine causing the population to not get vaccinated anymore. 


# The CMI-PB database
The CMI-RB project and associated database is making data on the immune response to Pertussis boost available to the scientific community. 

We will see what type of data CMI-PB make available here in a moment. 

First we need to be able to read data from CMI-PB. It, like many large advanced databases, returns JSON format. 

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
```

```{r}
head(subject, 3)
```


##Q4. How many aP and wP infancy vaccinated subjects are in the dataset? 

There are 47 aP infancy vaccinated subjects. There are 49 wP infancy vaccinated subjects. 

```{r}
table(subject$infancy_vac)
```

##Q5. How many Male and Female subjects/patients are in the dataset?

There are 66 female subjects and 30 male subjects in the dataset.

```{r}
table(subject$biological_sex)
```

```{r}
View(subject)
```


##Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

American Indian/Alaska Native: 0 Females, 1 Male
Asian: 18 Females, 9 Males
Black or African American: 2 Females, 0 Male
More Than One Race: 8 Females, 2 Males
Native Hawaiian or Other Pacific Islander: 1 Female, 1 Male
Unknown or Not Reported: 10 Females, 4 Males
White: 27 Females, 13 Males

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(lubridate)
```

##What is today’s date?
```{r}
today()
```

##How many days have passed since new year 2000?
```{r}
today() - ymd("2000-01-01")
```

##What is this in years?

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```


```{r}
# Use todays date to calculate age in days
subject$age <- time_length( today() - ymd(subject$year_of_birth), "years")
```


##Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

The average age of wP individuals is 36.

The average age of aP individuals is 25.

Based off the p-value (2.2e-16) obtained from a T-test it demonstrates that they are are significantly different. 

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary(ap$age))
```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")

round( summary(wp$age))
```

```{r}
t.test(ap$age, wp$age)
```

##Q8. Determine the age of all individuals at time of boost?

The ages of all individuals at time of boost are 30, 51, 33, 28, 25 and 28. 

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


What is the average age across all subjects?

```{r}
summary(subject$age)
```


Histogram

```{r}
ggplot(subject) + aes(age)+ geom_histogram()
```

And faceted by wP and aP


```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) 
```

```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

##Q9. With the help of a faceted boxplot or histogram, do you think these two groups are significantly different? 

Yes, based off the p-value (1.316045e-16) both groups are significantly different. 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
```

```{r}
head(specimen)
```

We want to "join" the 'subject' and 'specimen' tables to have all the metadata we need for later analysis. We can use the dplyr '*_join()' functions for this task. 

##Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(specimen, subject)
dim(meta)
```


```{r}
titer <- read_json("http://cmi-pb.org/api/ab_titer", simplifyVector = TRUE)
```

```{r}
head(titer)
```

I need to link a.k.a join this ab titer data with our meta data to have all the information we need to do some actual analysis like comparing Ab levels at different times after boost, comparing ap vs wP responses over time etc. 

##Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.
```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```


##Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

IgE: 6698
IgG: 1413
IgG1: 6141
IgG2: 6141
IgG3: 6141
IgG4: 6141
```{r}
table(abdata$isotype)
```

```{r}
ggplot(abdata) + aes(MFI, antigen) + geom_boxplot()
```


```{r}
ggplot(abdata) + aes(MFI_normalised, antigen) + geom_boxplot()
```


##Q12. What do you notice about the number of visit 8 specimens compared to other visits?

I notice that the number of visit 8 specimens is much lower compared to the other visits.

```{r}
table(abdata$visit)
```


```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```


##Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot()
```

We can break this up by visit (a quick proxy for time since boost).

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)
```

##Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?
Antigen FIM2/3 shows the greatest difference over time. Antigen FIM2/3 is different over time because it might have been used for the development of the vaccine. 


```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

##Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).

```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

















